On The Subspace Learning for Network Attack Detection

### 2019-06-14 docs/thesis/chapters/2_eigen_similarity/eigen_similarity.tex
after fix bold notation

In this work, scalars are denoted by italic letters ($a, b, A, B, \alpha, \beta$), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matrix, and the operator $\sim$ that sorts the elements of a vector in ascending order.

@@

%TODO - recreate the figure and change the concepts
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:2.01}
\end{figure}

@@

By modeling the data set as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{U}), artifact (\textbf{A}) and malicious traffic (\textbf{N}), according to the following expression:

\begin{equation}\label{eq:2.01}
	\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{A}^{(q)} + \boldsymbol{N}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time aggregation of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns, where each row represents a communication port, and each column represents time bins of a defined size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number packets that appears at $n$-th minute for the port $m$, during the $q$-th time frame.

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's ordinary operations, such as when performing access to web pages, as well when there is the traffic required to domain name resolution. Figure \ref{fig:2.03} depicts an example of the legitimate traffic simulated during the experiments.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:2.03}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as artifact $\boldsymbol{A}^{(q)}$. The acquisition of logical IP network address (DHCP) is an example of artifact, where independently of any user operation, the machine receives an IP address, since it is configured to automatically perform a DHCP address request. Figure \ref{fig:2.04} depicts an example of artifact in a network traffic, by means of traffic to ports 67 and 68. The traffic coming from a malicious activity, i.e. port scanning, synflood or fraggle attacks, is represented by the matrix $\boldsymbol{N}^{(q)}$.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2.04}
\end{figure}

We define that if the $\#\boldsymbol{X}^{(q)} ≠ 0$, then there is malicious traffic in the evaluated time frame $q$, on the other hand, if the $\#\boldsymbol{X}^{(q)} = 0$, then there is no malicious traffic. We show how to detect the $\#\boldsymbol{X}^{(q)}$, given only the matrix $\boldsymbol{X}^{(q)}$, in order to identify malicious traffic.

@@

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2.05}
\end{figure}

@@

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2.06}
\end{figure}

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2.07}
\end{figure}

@@

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:2.08}
\end{figure}

@@

With the data grouped into $Q$ time frames, the framework considers the time variations of the matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M\times{N}}$, with $q = 1, \ldots, Q$, in order to detect the attack.

@@

Therefore, to detect flood attacks, it is necessary to calculate the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ of the zero mean samples given by

\begin{equation}
\label{eq:2.02}
\boldsymbol{y}_{m}^{(q)} = \boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}.
\end{equation}

The set of obtained vectors $\boldsymbol{y}_{m}^{(q)}$ composes the zero mean matrix $\boldsymbol{Y}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ can be calculated as follows

\begin{equation}\label{eq:2.03}
\boldsymbol{\hat{R}}_{yy}^{(q)} = \frac{1}{N}\boldsymbol{Y}^{(q)}\boldsymbol{Y}^{(q)^{\rm T}}.
\end{equation}

For the detection of the port scan attack, the main components are not dominated by the variables with large variance. Moreover, the port scan traffic presents a highly correlated network traffic between the monitored ports. In order to exploit such structure, we compute the sample covariance $\boldsymbol{\hat{R}}_{zz}^{(q)}$ whose variables have zero mean and unitary standard deviation as follows

\begin{equation}
\label{eq:2.04}
\boldsymbol{z}_{m}^{(q)} = \frac{\boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}}{\boldsymbol{\sigma}_{m}^{(q)}}.
\end{equation}

The set of vectors $\boldsymbol{z}_{m}^{(q)}$ composes the matrix $\boldsymbol{Z}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{zz}^{(q)}$ can be calculated via

\begin{equation}\label{eq:2.05}
\boldsymbol{\hat{R}}_{zz}^{(q)} = \frac{1}{N}\boldsymbol{Z}^{(q)}\boldsymbol{Z}^{(q)^{\rm T}}.
\end{equation}

Once the $\boldsymbol{\hat{R}}_{yy}^{(q)}$ and $\boldsymbol{\hat{R}}_{zz}^{(q)}$ have been obtained for flood and port scan detection, respectively, and since the next steps are the same for both sample covariance matrices, we refer to $\boldsymbol{\hat{R}}_{yy}$ and $\boldsymbol{\hat{R}}_{zz}$ as a matrix $\boldsymbol{\hat{R}}$. Therefore, the following step of the algorithm is the eigenvalue decomposition (EVD), calculated according to (\ref{eq:2.06}), in order to obtain the vector of eigenvalues $\boldsymbol{e}^{(q)}$ associated with each matrix, according to (\ref{eq:2.06}).

\begin{equation}\label{eq:2.06}
\boldsymbol{\hat{R}}^{(q)} = \boldsymbol{V}^{(q)}\boldsymbol{\Lambda}^{(q)}\boldsymbol{V}^{(q)^{\rm T}},
\end{equation}

\begin{equation}\label{eq:2.060}
\boldsymbol{e}^{(q)} = \rm diag(\boldsymbol{\Lambda}^{(q)}),
\end{equation}

@@

The concatenation of the eigenvalues vector $\boldsymbol{e}^{(q)}$ for $q = 1, \ldots, Q$ is represented by

\begin{equation}\label{eq:2.07}
\boldsymbol{E} =
\begin{bmatrix}
  \lambda_1^{(1)} & \lambda_1^{(2)} & \lambda_1^{(3)} & \cdots & \lambda_1^{(Q)} \\
  \lambda_2^{(1)} & \lambda_2^{(2)} & \lambda_2^{(3)} & \cdots & \lambda_2^{(Q)} \\
  \lambda_3^{(1)} & \lambda_3^{(2)} & \lambda_3^{(3)} & \cdots & \lambda_3^{(Q)} \\
  \vdots & \vdots & \ddots & \vdots  \\
  \lambda_m^{(1)} & \lambda_m^{(2)} & \lambda_m^{(3)} & \cdots & \lambda_m^{(Q)} \\
\end{bmatrix}.
\end{equation}

Note that since $\lambda_1^{(q)} > \lambda_2^{(q)} > \lambda_3^{(q)} > \cdots > \lambda_{m-1}^{(q)} > \lambda_m^{(q)}$, then the first line of the matrix $\boldsymbol{E}$ contains the largest eigenvalues of each $q$-th time frame, which is the Greatest Eigenvalue Time
Vector (GETV) \cite{tenorio2013greatest}, denoted as

\begin{equation}\label{eq:2.08}
\boldsymbol{e}_{\rm max} = [ \lambda_1^{(1)}, \lambda_1^{(2)} ... \lambda_1^{(Q)}]
\end{equation}

@@

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\boldsymbol{e}^{(q)}$. However, the goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, the proposed approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\boldsymbol{e}_{\rm max}$ is sorted in descending order, producing $\sim\boldsymbol{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\boldsymbol{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\boldsymbol{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to the Appendix \ref{apx:a_mos}.

@@

After applying the MOS schemes to the vector $\sim\boldsymbol{e}_{\rm max}$, we obtain the estimate of the $\#\boldsymbol{X}$. For instance, in the case of fraggle, synflood and ports can, if $\hat{d} = 1$, , then $\#\boldsymbol{X} = 1$, which means that during the during the $Q$ time frames one attack is present. However, if $\hat{d} = 0$, then $\#\boldsymbol{X} = 0$, and this means that none of these attacks are present. Note that $\hat{d}$ can be greater than 1, indicating the presence of more than one attack.

@@

The largest eigenvalue analysis for estimating the $q$-th time frames that are under attack can be expressed according to Algorithm \ref{alg:2.01}, where $\rm{\boldsymbol{\hat{q}}}_{\rm max} \in \mathbb{R}^{\hat{d}}$ denotes a vector of the $q$-th time frames under attack, which is the $q$-th indexes corresponding to the $\hat{d}$ largest eigenvalues of $\boldsymbol{e}_{\rm max}$. Algorithm \ref{alg:2.01} initially identifies the largest value of $\boldsymbol{e}_{\rm max}$, according to Line 3 of Algorithm \ref{alg:2.01}, and its correspondent index, according to Line 7. Subsequently, the largest value is removed of $\boldsymbol{e}_{\rm max}$, according to Line 11 of Algorithm \ref{alg:2.01}, and a new iteration is performed until $\boldsymbol{e}_{\rm max} = []$.

\begin{algorithm}
	\label{alg:2.01}
	\SetAlgoLined
	\KwResult{$\rm{\boldsymbol{\hat{q}}}_{\rm max}$}
	Given $f = 1$\;
	\While{$f < \hat{d}$}{
		$q_{\rm value} = \operatorname*{argmax}_{\lambda}  \hspace{1 mm} \boldsymbol{e}_{\rm max}$\;
		$i = 1$\;
    	\While{$i < Q$}{
    		\If{$\boldsymbol{e}_{\rm max}^{(i)} == q_{\rm value}$}{
    			$\boldsymbol{\hat{q}}_{\rm max}^{(f)} = i$\;
    		}
    		$i = i + 1$\;
    	}
    	$\boldsymbol{e}_{\rm max} \rightarrow \boldsymbol{\hat{q}}_{\rm max}^{(f)}$\;
		$f = f + 1$\;
	}
	\caption{Detection of Time Frames Under Attack}
\end{algorithm}

After the estimation of the $\rm{\boldsymbol{\hat{q}}}_{\rm max}$ time frames under attack, it is necessary to obtain more details of the detected Attacks, such as the $n$-th minutes when the attacks happened and the $m$-th network ports that were attacked. To deal with this problem, the adoption of a similarity analysis between legitimate traffic and the traffic of time frames estimated as under attack is evaluated, analyzing the effectiveness of cosine similarity to highlight abnormalities inserted by network traffic attacks.

@@

Cosine similarity calculates the cosine of the angle between two vectors, which represents the similarity of values between the selected vectors. Therefore, cosine similarity can be used to evaluate the variation of the most significant eigenvectors of $\boldsymbol{V}^{(q)}$ against the most significant eigenvectors of time frame detected as under attack, to analyze similarity changes into the most significant eigenvectors caused by the insertion of anomalous traffic \cite{Lee2013}.

@@

\begin{equation}\label{eq:2.11}
s_n = \frac{\abs{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(n)}^{(q)}}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(n)}^{(q)}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\boldsymbol{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, according the attack detection described by Algorithm \ref{alg:2.01}, and $\boldsymbol{v}_{(n)}^{(q)}$ is the most significant eigenvectors obtained after append each target $n$-th minute of traffic that needs the identification of flood and port scan attacks.

The most significant eigenvector $\boldsymbol{v}^{(q)}$, of a time frame $q$ without attack, can be computed from (\ref{eq:2.06}) and selected according to the eigenvector of the largest eigenvalue $\lambda_1^{(q)}$, which is the principal component of the selected time frame $q$. The same calculation shall be performed in order to obtain the target eigenvectors $\boldsymbol{v}_{(n)}^{(q)}$, calculated from the time frame without attack appended by selected minutes of a time frame estimated as under attack.

Therefore, the reference eigenvectors $\boldsymbol{v}^{(q)}$ is calculated from the traffic without attack, in a time frame $q$ composed of $Q$ minutes of legitimate network traffic, estimated as normal by Algorithm \ref{alg:2.01}. For the detailed attack identification, each $\boldsymbol{x}^{(\hat{q})}_{(n)}$ vector of each $n$-th minutes of the estimated $\rm{\boldsymbol{\hat{q}}}_{\rm max}$ time frames shall be individually appended into $\boldsymbol{X}^{(q)}$, as represented by

\begin{equation}\label{eq:2.12}
\boldsymbol{X}_{n} = \{\boldsymbol{X}^{(q)} | \boldsymbol{x}^{(\hat{q})}_{(n)}\}.
\end{equation}

Subsequently, the resultant $\boldsymbol{X}_{(n)}$ is used to obtain $\boldsymbol{v}_{(n)}^{(q)}$, through (\ref{eq:2.06}), for calculating the similarity degree $s_n$, ranging from 0 to 1, for each $n$-th minute. The $s_n$ denotes the absolute similarity degree of the $n$-th minute in comparison to a well-known traffic without attack, detected through MOS schemes and eigenvalue analysis.

@@

The incremental approach for eigen similarity analysis is based on the incremental appending of network traffic into $\boldsymbol{X}^{(q)}$, where the first evaluation is based on (\ref{eq:2.12}) and the subsequent evaluations are based on (\ref{eq:2.13}), that denotes the appending of the $n$-th minute $\boldsymbol{x}^{(\hat{q})}_{(n)}$ into $\boldsymbol{X_n}$. The incremental approach repeats the Equation \ref{eq:2.13} while $n \leq N$ and compute $\boldsymbol{v}^{(q)}$, $\boldsymbol{v}_{(n)}^{(q)}$ and the Equation \ref{eq:2.11} for each increment.

\begin{equation}\label{eq:2.13}
\boldsymbol{X}_{n} = \{\boldsymbol{X}_{n} | \boldsymbol{x}^{(\hat{q})}_{(n)}\},
\end{equation}

Figure \ref{fig:2.08} illustrates the network traffic selection for the incremental approach of eigen similarity analysis, where the $\boldsymbol{X}^{(1)}$ is chosen as reference for similarity analysis of the $m$-th minutes of the time frame $q=3$, where one network attack was previously detected by Algorithm \ref{alg:2.01}.

%TODO - fix figure that shows s_n > l where should the inverse
\begin{figure}[h!]
     \includegraphics[width=15cm]{figures/incremental.eps}
     \caption{Traffic selection for incremental approach.}
     \label{fig:2.08}
\end{figure}

For the scenario depicted by Figure \ref{fig:2.08}, the eigen similarity analysis starts at $\boldsymbol{x}^{(3)}_{(41)}$ and is incrementally performed until $\boldsymbol{x}^{(3)}_{(60)}$, in order to calculate the $s_n$. We assume that $s_n < l$ means an attack identification, according the anomaly on similarity of $s_n$ compared to a defined limiar $l$.

Therefore, after obtaining the most significant eigenvector $\boldsymbol{v}^{(q)}$ and the target eigenvectors $\boldsymbol{v}_{(n)}^{(q)}$ for eigen similarity analysis, the $s_n$ is calculated according to (\ref{eq:2.11}). If $s_n = 1$, then the two eigenvectors are completely similar and no anomaly is detected. Smaller values of $s_n$ mean less similarity and can indicate an anomaly if $s_n < l$, what denotes that a network attack is identified during the $n$-th minute.

>>>>

In this work, scalars are denoted by italic letters ($a, b, A, B, \alpha, \beta$), vectors by lowercase bold letters ($\pmb{a}, \pmb{b}$), matrices by uppercase bold letters ($\pmb{A}, \pmb{B}$), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix $\pmb{A}$. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matrix, and the operator $\sim$ that sorts the elements of a vector in ascending order.

@@

%TODO - recreate the figure and change the concepts
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:2.01}
\end{figure}

@@

By modeling the data set as a signal superposition, the network traffic ($\pmb{X}$) can be characterized as a mixture of three components: legitimate traffic ($\pmb{U}$), artifact ($\pmb{A}$) and malicious traffic ($\pmb{N}$), according to the following expression:

\begin{equation}\label{eq:2.01}
	\pmb{X}^{(q)} = \pmb{U}^{(q)} + \pmb{A}^{(q)} + \pmb{N}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time aggregation of network traffic. The matrix $\pmb{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns, where each row represents a communication port, and each column represents time bins of a defined size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number packets that appears at $n$-th minute for the port $m$, during the $q$-th time frame.

The legitimate traffic $\pmb{U}^{(q)}$ is characterized by the traffic from user's ordinary operations, such as when performing access to web pages, as well when there is the traffic required to domain name resolution. Figure \ref{fig:2.03} depicts an example of the legitimate traffic simulated during the experiments.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:2.03}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as artifact $\pmb{A}^{(q)}$. The acquisition of logical IP network address (DHCP) is an example of artifact, where independently of any user operation, the machine receives an IP address, since it is configured to automatically perform a DHCP address request. Figure \ref{fig:2.04} depicts an example of artifact in a network traffic, by means of traffic to ports 67 and 68. The traffic coming from a malicious activity, i.e. port scanning, synflood or fraggle attacks, is represented by the matrix $\pmb{N}^{(q)}$.

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2.04}
\end{figure}

We define that if the $\#\pmb{X}^{(q)} ≠ 0$, then there is malicious traffic in the evaluated time frame $q$, on the other hand, if the $\#\pmb{X}^{(q)} = 0$, then there is no malicious traffic. We show how to detect the $\#\pmb{X}^{(q)}$, given only the matrix $\pmb{X}^{(q)}$, in order to identify malicious traffic.

@@

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2.05}
\end{figure}

@@

% TODO - Replace
\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2.06}
\end{figure}

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/ch2/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2.07}
\end{figure}

@@

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/ch2/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:2.08}
\end{figure}

@@

With the data grouped into $Q$ time frames, the framework considers the time variations of the matrix $\pmb{X}^{(q)} \in \mathbb{R}^{M\times{N}}$, with $q = 1, \ldots, Q$, in order to detect the attack.

@@

Therefore, to detect flood attacks, it is necessary to calculate the sample covariance matrix $\hat{\pmb{R}}_{yy}^{(q)}$ of the zero mean samples given by

\begin{equation}
\label{eq:2.02}
\pmb{y}_{m}^{(q)} = \pmb{x}_{m}^{(q)} - \bar{\pmb{x}}_{m}^{(q)}.
\end{equation}

The set of obtained vectors $\pmb{y}_{m}^{(q)}$ composes the zero mean matrix $\pmb{Y}^{(q)}$, then the sample covariance matrix $\hat{\pmb{R}}_{yy}^{(q)}$ can be calculated as follows

\begin{equation}\label{eq:2.03}
\hat{\pmb{R}}_{yy}^{(q)} = \frac{1}{N}\pmb{Y}^{(q)}\pmb{Y}^{(q)^{\rm T}}.
\end{equation}

For the detection of the port scan attack, the main components are not dominated by the variables with large variance. Moreover, the port scan traffic presents a highly correlated network traffic between the monitored ports. In order to exploit such structure, we compute the sample covariance $\hat{\pmb{R}}_{zz}^{(q)}$ whose variables have zero mean and unitary standard deviation as follows

\begin{equation}
\label{eq:2.04}
\pmb{z}_{m}^{(q)} = \frac{\pmb{x}_{m}^{(q)} - \bar{\pmb{x}}_{m}^{(q)}}{\pmb{\sigma}_{m}^{(q)}}.
\end{equation}

The set of vectors $\pmb{z}_{m}^{(q)}$ composes the matrix $\pmb{Z}^{(q)}$, then the sample covariance matrix $\hat{\pmb{R}}_{zz}^{(q)}$ can be calculated via

\begin{equation}\label{eq:2.05}
\hat{\pmb{R}}_{zz}^{(q)} = \frac{1}{N}\pmb{Z}^{(q)}\pmb{Z}^{(q)^{\rm T}}.
\end{equation}

Once the $\hat{\pmb{R}}_{yy}^{(q)}$ and $\hat{\pmb{R}}_{zz}^{(q)}$ have been obtained for flood and port scan detection, respectively, and since the next steps are the same for both sample covariance matrices, we refer to $\hat{\pmb{R}}_{yy}$ and $\hat{\pmb{R}}_{zz}$ as a matrix $\hat{\pmb{R}}$. Therefore, the following step of the algorithm is the eigenvalue decomposition (EVD), calculated according to (\ref{eq:2.06}), in order to obtain the vector of eigenvalues $\pmb{e}^{(q)}$ associated with each matrix, according to (\ref{eq:2.06}).

\begin{equation}\label{eq:2.06}
\hat{\pmb{R}}^{(q)} = \pmb{V}^{(q)}\pmb{\Lambda}^{(q)}\pmb{V}^{(q)^{\rm T}},
\end{equation}

\begin{equation}\label{eq:2.060}
\pmb{e}^{(q)} = \rm diag(\pmb{\Lambda}^{(q)}),
\end{equation}

@@

The concatenation of the eigenvalues vector $\pmb{e}^{(q)}$ for $q = 1, \ldots, Q$ is represented by

\begin{equation}\label{eq:2.07}
\pmb{E} =
\begin{bmatrix}
  \lambda_1^{(1)} & \lambda_1^{(2)} & \lambda_1^{(3)} & \cdots & \lambda_1^{(Q)} \\
  \lambda_2^{(1)} & \lambda_2^{(2)} & \lambda_2^{(3)} & \cdots & \lambda_2^{(Q)} \\
  \lambda_3^{(1)} & \lambda_3^{(2)} & \lambda_3^{(3)} & \cdots & \lambda_3^{(Q)} \\
  \vdots & \vdots & \ddots & \vdots  \\
  \lambda_m^{(1)} & \lambda_m^{(2)} & \lambda_m^{(3)} & \cdots & \lambda_m^{(Q)} \\
\end{bmatrix}.
\end{equation}

Note that since $\lambda_1^{(q)} > \lambda_2^{(q)} > \lambda_3^{(q)} > \cdots > \lambda_{m-1}^{(q)} > \lambda_m^{(q)}$, then the first line of the matrix $\pmb{E}$ contains the largest eigenvalues of each $q$-th time frame, which is the Greatest Eigenvalue Time
Vector (GETV) \cite{tenorio2013greatest}, denoted as

\begin{equation}\label{eq:2.08}
\pmb{e}_{\rm max} = [ \lambda_1^{(1)}, \lambda_1^{(2)} ... \lambda_1^{(Q)}]
\end{equation}

@@

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\pmb{e}^{(q)}$. However, the goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, the proposed approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\pmb{e}_{\rm max}$ is sorted in descending order, producing $\sim\pmb{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\pmb{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\pmb{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to the Appendix \ref{apx:a_mos}.

@@

After applying the MOS schemes to the vector $\sim\pmb{e}_{\rm max}$, we obtain the estimate of the $\#\pmb{X}$. For instance, in the case of fraggle, synflood and ports can, if $\hat{d} = 1$, , then $\#\pmb{X} = 1$, which means that during the during the $Q$ time frames one attack is present. However, if $\hat{d} = 0$, then $\#\pmb{X} = 0$, and this means that none of these attacks are present. Note that $\hat{d}$ can be greater than 1, indicating the presence of more than one attack.

@@

The largest eigenvalue analysis for estimating the $q$-th time frames that are under attack can be expressed according to Algorithm \ref{alg:2.01}, where $\rm{\hat{\pmb{q}}}_{\rm max} \in \mathbb{R}^{\hat{d}}$ denotes a vector of the $q$-th time frames under attack, which is the $q$-th indexes corresponding to the $\hat{d}$ largest eigenvalues of $\pmb{e}_{\rm max}$. Algorithm \ref{alg:2.01} initially identifies the largest value of $\pmb{e}_{\rm max}$, according to Line 3 of Algorithm \ref{alg:2.01}, and its correspondent index, according to Line 7. Subsequently, the largest value is removed of $\pmb{e}_{\rm max}$, according to Line 11 of Algorithm \ref{alg:2.01}, and a new iteration is performed until $\pmb{e}_{\rm max} = []$.

\begin{algorithm}
	\label{alg:2.01}
	\SetAlgoLined
	\KwResult{$\rm{\hat{\pmb{q}}}_{\rm max}$}
	Given $f = 1$\;
	\While{$f < \hat{d}$}{
		$q_{\rm value} = \operatorname*{argmax}_{\lambda}  \hspace{1 mm} \pmb{e}_{\rm max}$\;
		$i = 1$\;
    	\While{$i < Q$}{
    		\If{$\pmb{e}_{\rm max}^{(i)} == q_{\rm value}$}{
    			$\hat{\pmb{q}}_{\rm max}^{(f)} = i$\;
    		}
    		$i = i + 1$\;
    	}
    	$\pmb{e}_{\rm max} \rightarrow \hat{\pmb{q}}_{\rm max}^{(f)}$\;
		$f = f + 1$\;
	}
	\caption{Detection of Time Frames Under Attack}
\end{algorithm}

After the estimation of the $\rm{\hat{\pmb{q}}}_{\rm max}$ time frames under attack, it is necessary to obtain more details of the detected Attacks, such as the $n$-th minutes when the attacks happened and the $m$-th network ports that were attacked. To deal with this problem, the adoption of a similarity analysis between legitimate traffic and the traffic of time frames estimated as under attack is evaluated, analyzing the effectiveness of cosine similarity to highlight abnormalities inserted by network traffic attacks.

@@

Cosine similarity calculates the cosine of the angle between two vectors, which represents the similarity of values between the selected vectors. Therefore, cosine similarity can be used to evaluate the variation of the most significant eigenvectors of $\pmb{V}^{(q)}$ against the most significant eigenvectors of time frame detected as under attack, to analyze similarity changes into the most significant eigenvectors caused by the insertion of anomalous traffic \cite{Lee2013}.

@@

\begin{equation}\label{eq:2.11}
s_n = \frac{\abs{\pmb{v}^{(q)} \cdot \pmb{v}_{(n)}^{(q)}}}{\norm{\pmb{v}^{(q)}}\norm{\pmb{v}_{(n)}^{(q)}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\pmb{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, according the attack detection described by Algorithm \ref{alg:2.01}, and $\pmb{v}_{(n)}^{(q)}$ is the most significant eigenvectors obtained after append each target $n$-th minute of traffic that needs the identification of flood and port scan attacks.

The most significant eigenvector $\pmb{v}^{(q)}$, of a time frame $q$ without attack, can be computed from (\ref{eq:2.06}) and selected according to the eigenvector of the largest eigenvalue $\lambda_1^{(q)}$, which is the principal component of the selected time frame $q$. The same calculation shall be performed in order to obtain the target eigenvectors $\pmb{v}_{(n)}^{(q)}$, calculated from the time frame without attack appended by selected minutes of a time frame estimated as under attack.

Therefore, the reference eigenvectors $\pmb{v}^{(q)}$ is calculated from the traffic without attack, in a time frame $q$ composed of $Q$ minutes of legitimate network traffic, estimated as normal by Algorithm \ref{alg:2.01}. For the detailed attack identification, each $\pmb{x}^{(\hat{q})}_{(n)}$ vector of each $n$-th minutes of the estimated $\rm{\hat{\pmb{q}}}_{\rm max}$ time frames shall be individually appended into $\pmb{X}^{(q)}$, as represented by

\begin{equation}\label{eq:2.12}
\pmb{X}_{n} = \{\pmb{X}^{(q)} | \pmb{x}^{(\hat{q})}_{(n)}\}.
\end{equation}

Subsequently, the resultant $\pmb{X}_{(n)}$ is used to obtain $\pmb{v}_{(n)}^{(q)}$, through (\ref{eq:2.06}), for calculating the similarity degree $s_n$, ranging from 0 to 1, for each $n$-th minute. The $s_n$ denotes the absolute similarity degree of the $n$-th minute in comparison to a well-known traffic without attack, detected through MOS schemes and eigenvalue analysis.

@@

The incremental approach for eigen similarity analysis is based on the incremental appending of network traffic into $\pmb{X}^{(q)}$, where the first evaluation is based on (\ref{eq:2.12}) and the subsequent evaluations are based on (\ref{eq:2.13}), that denotes the appending of the $n$-th minute $\pmb{x}^{(\hat{q})}_{(n)}$ into $\pmb{X_n}$. The incremental approach repeats the Equation \ref{eq:2.13} while $n \leq N$ and compute $\pmb{v}^{(q)}$, $\pmb{v}_{(n)}^{(q)}$ and the Equation \ref{eq:2.11} for each increment.

\begin{equation}\label{eq:2.13}
\pmb{X}_{n} = \{\pmb{X}_{n} | \pmb{x}^{(\hat{q})}_{(n)}\},
\end{equation}

Figure \ref{fig:2.08} illustrates the network traffic selection for the incremental approach of eigen similarity analysis, where the $\pmb{X}^{(1)}$ is chosen as reference for similarity analysis of the $m$-th minutes of the time frame $q=3$, where one network attack was previously detected by Algorithm \ref{alg:2.01}.

%TODO - fix figure that shows s_n > l where should the inverse
\begin{figure}[h!]
     \includegraphics[width=15cm]{figures/ch2/incremental.eps}
     \caption{Traffic selection for incremental approach.}
     \label{fig:2.08}
\end{figure}

For the scenario depicted by Figure \ref{fig:2.08}, the eigen similarity analysis starts at $\pmb{x}^{(3)}_{(41)}$ and is incrementally performed until $\pmb{x}^{(3)}_{(60)}$, in order to calculate the $s_n$. We assume that $s_n < l$ means an attack identification, according the anomaly on similarity of $s_n$ compared to a defined limiar $l$.

Therefore, after obtaining the most significant eigenvector $\pmb{v}^{(q)}$ and the target eigenvectors $\pmb{v}_{(n)}^{(q)}$ for eigen similarity analysis, the $s_n$ is calculated according to (\ref{eq:2.11}). If $s_n = 1$, then the two eigenvectors are completely similar and no anomaly is detected. Smaller values of $s_n$ mean less similarity and can indicate an anomaly if $s_n < l$, what denotes that a network attack is identified during the $n$-th minute.
